---
title: "Final Visualizations and Analysis"
author: "ElectriGrid"
format: html
---

The following notebook contains code to create visualizations for:

1.  Equality ratios (bar plot)

2.  Boxplots comparing electric grid variables for disadvantaged and non-disadvantaged communities

Adaptions of these visualizations were used in the final public presentation for ElectriGrid, and also provide an exploration of relationships between sociodemographic variables and the electrid grid in California. 

```{r}
#| message: false 
# import necessary libraries
library(tidyverse)
library(patchwork)
```

## Equity Ratios

```{r}
#| message: false
# import data
ratios_raw <- read_csv("ratios.csv")
```

```{r}
head(ratios_raw)
```

```{r}
# keep specific ratios of interest (remaining capacity)
ratios <- ratios_raw %>% 
  select(group, avg_DER_remain_generation_hh, avg_DER_remain_load_hh, avg_DER_remain_pv_hh) %>% 
  pivot_longer(cols = c(avg_DER_remain_generation_hh,avg_DER_remain_pv_hh), names_to = "variable", values_to = "ratio")
```

```{r}
# save colors
blue <- "#bfd8e9"
solar <- "#E6C75E"
gen <- "#A8CB92"
```

```{r}
# plot
ggplot(ratios, aes(x = group, y = ratio, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge",
           color = "white") +
  geom_hline(yintercept = 1, linetype = "dashed", linewidth = 1.5,
             color = blue) +
  scale_fill_manual(values = c(gen, solar), 
                    labels = c("Total", "Solar")) +
  scale_x_discrete(labels = c("Asian_Pct" = "Asian", "Black_Pct" = "Black",
                              "Hispanic_Pct" = "Hispanic", "NatAmeri_Pct" = "Native American",
                              "OtherMulti_Pct" = "Other/Multi", "White_Pct" = "White")) +
  labs(fill = " ",
       y = "Equity Ratio", x = "Racial/Ethnic Group") +
  theme_minimal()

# ggsave("ratios_plot.png", height = 4, width = 8)
```

**Conclusion:** Hispanic communities appear to be under-represented in both total and solar hosting capacity in California.

## Distribution Boxplots

```{r}
#| message: false
df <- read_csv("dis_hosting_075.csv")
```

Creating 6 boxplots: one for each generation capacity variable that we calculated.

```{r}
# save variable names to iterate over
cols = c('avg_DER_remain_pv_opflex_hh', 'avg_DER_remain_generation_hh', 'avg_DER_remain_generation_opflex_hh', 'avg_DER_total_generation_hh', 'avg_DER_remain_load_hh', 'avg_DER_total_pv_hh')


# create a boxplot for each capacity variable, comparing the different DAC statuses

for (i in 1:length(cols)){ # iterate over each variable name in df


assign(paste0("plot_", cols[i]), # assign plot name based on generation variable
       
       # create basic boxplot
       ggplot(df, aes(y = as.factor(DAC_status), .data[[cols[i]]])) +
       geom_boxplot(outliers = FALSE) +
       labs(y = "DAC Status")
       + theme_light()) 

}
```

Assembling the 6 plots into one figure to view.

```{r}
#| warning: false
# using patchwork, assemble 6 plots into 1
p <- plot_avg_DER_remain_pv_opflex_hh + plot_avg_DER_remain_generation_hh +
plot_avg_DER_remain_generation_opflex_hh + plot_avg_DER_total_generation_hh + plot_avg_DER_remain_load_hh + plot_avg_DER_total_pv_hh +
  
  # add title
  plot_annotation(title = "Comparing distribution of electric grid variables (kW) between disadvantaged and\nnon-disadvantaged communities",
    subtitle = "Where DAC Status = 1 refers to CI Score being in the 75th percentile")

# save figure
ggsave("boxplots.png", plot = p, width = 9.5, height = 6, unit = "in")

# view
p
```

**Conclusion**: Disadvanataged communities have a lower median total hosting capacity but more capacity to host solar.