---
title: "Investigating the Association between Census Demographics and Hosting Capactiy"
---

# Load libraries & data
```{r}
library(tidyverse)
library(sfarrow)
library(tidycensus)
library(sf)
library(here)
library(tidymodels)
```

```{r}
census <- read_csv('census_final_tidy.csv') |> 
  janitor::clean_names() |> 
  select(-geometry)

census_area <- st_read("~/../../capstone/electrigrid/data/census/census_tracts_2025/tl_2025_06_tract.shp") |> 
  select(GEOID, ALAND) |> janitor::clean_names()

dac <- st_read_parquet("~/../../capstone/electrigrid/data/dis_adv_comm/calenviroscreen.parquet") %>% 
  mutate(tract = as.character(tract)) %>% 
  mutate(tract = str_pad(tract, width = 11, side = "left", pad = "0"))

dir <- here::here('~', '..', '..', 'capstone', 'electrigrid', 'data', 'census_hosting_capacity')

sce_census_map <- sfarrow::st_read_parquet(here(dir, "sce_census_map_new.parquet")) |> 
    janitor::clean_names() |> 
    select(geoid, zillow_tract_hh_count, starts_with('avg_der'), geometry)

pge_census_map <- sfarrow::st_read_parquet(here(dir, "pge_census_map_new.parquet")) |> 
    janitor::clean_names() |> 
    select(geoid, zillow_tract_hh_count, starts_with('avg_der'), geometry)

sdge_census_map <- sfarrow::st_read_parquet(here(dir, "sdge_census_map_new.parquet")) |> 
    janitor::clean_names() |> 
    select(geoid, zillow_tract_hh_count, starts_with('avg_der'), geometry)
```

# Clean data

```{r}
# combine all utilities
all_census_map <- bind_rows(sce_census_map, pge_census_map, sdge_census_map) |> 
    # set minimum zillow count
    mutate(zillow_tract_hh_count = pmax(zillow_tract_hh_count, 1))
```

```{r}
# iou service areas overlap, must handle overlapped tracts
overlap_tracts <- all_census_map |> # 1620 overlapped tracts
  group_by(geoid) |> 
  filter(n() > 1) |> 
  pull(geoid) |> unique()

# overlap_rows <- all_census_map |> 
#   filter(geoid %in% overlap_tracts) |> arrange(geoid)

# fix overlapped tracts with weighted mean
all_census_map <- all_census_map |>
  group_by(geoid) |>
  summarise(
    across(starts_with("avg_der"), ~ weighted.mean(.x, w = zillow_tract_hh_count, na.rm = TRUE)),
    zillow_tract_hh_count = sum(zillow_tract_hh_count)
  ) |> ungroup()
```

```{r}
# need to handle outliers in combined zillow estimate
all_census_map |>
  arrange(desc(zillow_tract_hh_count)) |>
  select(geoid, zillow_tract_hh_count) |>
  st_drop_geometry() |> 
  head(20)

hist(log10(all_census_map$zillow_tract_hh_count))
```

The histogram shows our estimates are centered around 10^3 units, which, in the census data lines up with outliers above 10^4 should not be included.

```{r}
all_census_map_filtered <- all_census_map |> 
  filter(zillow_tract_hh_count <= 10000)
```

# Random Forest Revised with only DAC data

```{r}
features <- c(
  # Race/ethnicity (drop one to avoid perfect multicollinearity, drop White_Pct)
  "Hispanic_Pct", "Black_Pct", "Asian_Pct", "NatAmeri_Pct", "OtherMulti_Pct",
  
  # Socioeconomic
  "Poverty_Pctl", "HousingBurden_Pctl", "Education_Pctl", "CIscore_Pctl",
  "Unemployment_Pctl", "LinguisticIso_Pctl", "housing_density_hh_km2",
  
  # DAC status (binary)
  "is_dac"
)
```

```{r}
# prepare clean model data
model_data <- all_census_map_filtered |> 
  st_drop_geometry() |> 
  left_join(dac |> st_drop_geometry(), by = join_by(geoid == tract)) |> 
  mutate(
    is_dac = as.integer(CIscore_Pctl >= 75),
    diversity = 1 - ((White_Pct/100)^2 + (Hispanic_Pct/100)^2 + (Black_Pct/100)^2 + 
                 (Asian_Pct/100)^2 + (NatAmeri_Pct/100)^2 + (OtherMulti_Pct/100)^2),
    housing_density_hh_km2 = zillow_tract_hh_count / Shape_Area * 1e6
  ) |>
  select(geoid, starts_with('avg_der'), all_of(features), zillow_tract_hh_count) |> na.omit()
```

```{r}
# 2. recipe
rec <- recipe(avg_der_remain_pv_hh ~ ., data = model_data |> select(geoid, avg_der_remain_pv_hh, all_of(features), zillow_tract_hh_count)) |>
  update_role(geoid, new_role = "ID") |>
  update_role(zillow_tract_hh_count, new_role = "ID")  # used as weight, not predictor

# 3. model spec — mtry to tune
rf_spec <- rand_forest(mtry = tune(), trees = 500) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("regression")

# 4. workflow
wf <- workflow() |>
  add_recipe(rec) |>
  add_model(rf_spec)

# 5. 10-fold CV + tune mtry
cv <- vfold_cv(model_data, v = 10)
tune_res <- tune_grid(
  wf,
  resamples = cv,
  grid = tibble(mtry = c(2, 3, 11))  
)

# 6. fit final model with best mtry
best <- select_best(tune_res, metric = "rmse")
final_wf <- finalize_workflow(wf, best)
final_fit <- fit(final_wf, model_data)

# 7. results
collect_metrics(tune_res)
final_fit |> extract_fit_engine() |> vip::vip()
```

The resulting metrics & VIP show that this RF model has a low R-squared but still has meaningful feature importance, indicating that our features don't *predict* hosting capacity but are still *correlated*.

```{r}
# post model checks
model_data |>
  select(all_of(features)) |>
  cor(use = "complete.obs") |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "cor") |>
  filter(var1 < var2, abs(cor) > 0.5) |>
  arrange(desc(abs(cor)))
```

As the data we used is from CalEnviroScreen and some variables are suspected to be composites of others, the above cell investigates our features for collinearity. Collinear variables will be dropped before running a linear regression. We found that `Education_Pctl` and `CIscore_Pctl` is correlated with `Hispanic_Pct`, `Poverty_Pctl`, and `LinguisticIso_Pctl`. `is_dac` will also be dropped as it does not show in our VIP.

```{r}
lm_features <- c(
  "housing_density_hh_km2", "Hispanic_Pct", "Black_Pct", "Asian_Pct", "NatAmeri_Pct", 
  "OtherMulti_Pct", "Poverty_Pctl", "HousingBurden_Pctl", "Unemployment_Pctl", "LinguisticIso_Pctl"
)
```

```{r}
# Linear Regression (scaled & weighted) 
model_data_scaled <- model_data |> 
  mutate(across(all_of(lm_features), scale))

lm_fit <- lm(avg_der_remain_pv_hh ~ ., data = model_data_scaled |> select(avg_der_remain_pv_hh, all_of(lm_features)),
weights = model_data_scaled$zillow_tract_hh_count)

summary(lm_fit)
```

Before further investigation, we repeat this same process on another DER type.

```{r}
# 2. recipe
rec <- recipe(avg_der_remain_load_hh ~ ., data = model_data |> select(geoid, avg_der_remain_load_hh, all_of(features), zillow_tract_hh_count)) |>
  update_role(geoid, new_role = "ID") |>
  update_role(zillow_tract_hh_count, new_role = "ID")  # used as weight, not predictor

# 3. model spec — mtry to tune
rf_spec <- rand_forest(mtry = tune(), trees = 500) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("regression")

# 4. workflow
wf <- workflow() |>
  add_recipe(rec) |>
  add_model(rf_spec)

# 5. 10-fold CV + tune mtry
cv <- vfold_cv(model_data, v = 10)
tune_res <- tune_grid(
  wf,
  resamples = cv,
  grid = tibble(mtry = c(2, 3, 11))  
)

# 6. fit final model with best mtry
best <- select_best(tune_res, metric = "rmse")
final_wf <- finalize_workflow(wf, best)
final_fit <- fit(final_wf, model_data)

# 7. results
collect_metrics(tune_res)
final_fit |> extract_fit_engine() |> vip::vip()

lm_fit <- lm(avg_der_remain_load_hh ~ ., data = model_data_scaled |> select(avg_der_remain_load_hh, all_of(lm_features)),
weights = model_data_scaled$zillow_tract_hh_count)

summary(lm_fit)
```

Run our models again on remaining generation.

```{r}
# 2. recipe
rec <- recipe(avg_der_remain_generation_hh ~ ., data = model_data |> select(geoid, avg_der_remain_generation_hh, all_of(features), zillow_tract_hh_count)) |>
  update_role(geoid, new_role = "ID") |>
  update_role(zillow_tract_hh_count, new_role = "ID")  # used as weight, not predictor

# 3. model spec — mtry to tune
rf_spec <- rand_forest(mtry = tune(), trees = 500) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("regression")

# 4. workflow
wf <- workflow() |>
  add_recipe(rec) |>
  add_model(rf_spec)

# 5. 10-fold CV + tune mtry
cv <- vfold_cv(model_data, v = 10)
tune_res <- tune_grid(
  wf,
  resamples = cv,
  grid = tibble(mtry = c(2, 3, 11))  
)

# 6. fit final model with best mtry
best <- select_best(tune_res, metric = "rmse")
final_wf <- finalize_workflow(wf, best)
final_fit <- fit(final_wf, model_data)

# 7. results
collect_metrics(tune_res)
final_fit |> extract_fit_engine() |> vip::vip()

lm_fit <- lm(avg_der_remain_generation_hh ~ ., data = model_data_scaled |> select(avg_der_remain_generation_hh, all_of(lm_features)),
weights = model_data_scaled$zillow_tract_hh_count)

summary(lm_fit)
```

-------------- everything below is for previous iterations of a potential stats workflow, keeping it for now --------

# Prepare data for statistics

```{r}
# calenviroscreen join
der_cal <- left_join(all_census_map, dac %>% st_drop_geometry(), by = join_by(geoid == tract))
```

```{r}
der_cal |>
  select(geoid, avg_der_remain_pv_hh, White_Pct, Hispanic_Pct, Black_Pct, Asian_Pct) |>
  pivot_longer(cols = ends_with('_Pct'), names_to = 'race', values_to = 'race_pct') |>
  ggplot(aes(x = race_pct, y = avg_der_remain_pv_hh, color = race)) +
  geom_smooth(method = 'loess', se = TRUE) +
  facet_wrap(~race) +
  labs(x = 'Race/ethnicity (%)', y = 'Avg remaining PV hosting capacity (kW/HH)')
```

```{r}
der_cols <- der_cal |> st_drop_geometry() |> select(starts_with('avg_der')) |> names()

plots <- map(der_cols, \(der_type) {
  der_cal |>
    select(geoid, all_of(der_type), White_Pct, Hispanic_Pct, Black_Pct, Asian_Pct) |>
    pivot_longer(cols = ends_with('_Pct'), names_to = 'race', values_to = 'race_pct') |>
    mutate(race = str_remove(race, '_Pct')) |>
    ggplot(aes(x = race_pct, y = .data[[der_type]], color = race)) +
    geom_smooth(method = 'loess', se = TRUE) +
    guides(color = 'none') +
    facet_wrap(~race) +
    labs(x = 'Race/ethnicity (%)', y = der_type, title = der_type)
})

walk(plots, print)
```

```{r}
model <- lm(avg_der_total_pv_hh ~ White_Pct + Hispanic_Pct + Asian_Pct + Black_Pct, data = der_cal)
summary(model)
```

```{r}
der_long <- der_cal %>% 
  select(geoid, starts_with('avg_der'), 'White_Pct', 'Hispanic_Pct', 'Black_Pct', 'Asian_Pct') %>% 
  pivot_longer(cols=starts_with('avg_der'), names_to = 'der_type', values_to = 'hosting_cap') %>% 
  pivot_longer(cols=ends_with('Pct'), names_to = 'race', values_to = 'race_pct')
```

```{r}
ggplot(der_long %>% filter(hosting_cap >= 0), aes(x = race_pct, y = hosting_cap, color = race)) +
  geom_point() +
  facet_wrap(~der_type)
```



```{r}
# acs join
joined <- left_join(all_census_map, census, by = 'geoid')

joined_cleandem <- joined |> 
    filter(str_starts(variable, "B01003_") | str_starts(variable, "B02001")) |> 
    mutate(demographic = case_when(
        str_detect(label, "White") ~ "White",
        str_detect(label, "Black") ~ "Black",
        str_detect(label, "Total:!!Hispanic") ~ "Hispanic/Latino",
        str_detect(label, "Indian") ~ "American Indian/Alaska Native",
        str_detect(label, "Asian") ~ "Asian",
        str_detect(label, "Hawaiian") ~ "Native Hawaiian/Pacific Islander",
        str_detect(label, "Other Race") ~ "Other Race",
        str_detect(label, "Two or More") ~ "2+ Races",
        str_detect(label, "Total:!!Not") ~ "Not Hispanic/Latino",
        str_detect(label, "Estimate!!Total") ~ "Total Pop"
    )) |> 
    relocate(demographic, .after = label)

joined_wide <- joined_cleandem |>
  sf::st_drop_geometry() |> 
  pivot_wider(
    id_cols = c(geoid, starts_with('avg_der'), name),
    names_from = demographic,
    values_from = estimate) |> 
  janitor::clean_names() |> 
  mutate(
    diversity = 1 - (white*(white-1) + black*(black-1) + american_indian_alaska_native*(american_indian_alaska_native-1) + asian*(asian-1) + native_hawaiian_pacific_islander*(native_hawaiian_pacific_islander-1) + other_race*(other_race-1) + x2_races*(x2_races-1)) / (total_pop*(total_pop-1)))

percents <- joined_wide |> 
  mutate(
    white_pct = white / total_pop, 
    black_pct = black / total_pop, 
    native_indian_pct = american_indian_alaska_native / total_pop, 
    asian_pct = asian / total_pop,
    native_pacific_pct = native_hawaiian_pacific_islander / total_pop,
    other_pct = other_race / total_pop,
    x2_races_pct = x2_races / total_pop) |> 
      select(geoid, name, starts_with('avg_der'), ends_with('_pct'), diversity)
```

```{r}
model <- lm(avg_der_remain_pv_hh ~ white_pct + black_pct + asian_pct + x2_races_pct + diversity, data = percents)
summary(model)
```